---
title: "Interpretation of Economic Evaluation Results in HTA"
subtitle: "Interactive Session for Community Medicine Residents"
author: "RRC-HTA, AIIMS Bhopal"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
---

## Welcome {.unnumbered}

This interactive session will help you interpret economic evaluation results and make evidence-based policy decisions.

**Learning Objectives:**

1. Interpret ICER values and cost-effectiveness planes
2. Understand probabilistic sensitivity analysis results
3. Apply WTP thresholds to make reimbursement decisions
4. Identify key drivers of uncertainty using tornado diagrams
5. Interpret EVPI to judge research value
6. Assess budget impact and affordability
7. Identify common interpretation pitfalls

::: {.callout-tip}
## Quick Poll
Before we begin, what's your biggest challenge with economic evaluations?

- Understanding ICER calculations
- Interpreting uncertainty
- Applying thresholds
- Translating to policy decisions
:::

::: {.callout-note}
## Quick Recap: Key Concepts

These concepts are covered in detail by the earlier workshop sessions. A brief recap for context:

**QALY** (Quality-Adjusted Life Year) — One year of life in perfect health = 1 QALY. It captures both length and quality of life. For example, living 2 years at a utility of 0.7 = 1.4 QALYs. Utility weights typically come from patient-reported instruments such as EQ-5D.

**WTP Threshold** (Willingness-to-Pay) — The maximum amount a health system is willing to pay per QALY gained. It reflects opportunity cost: the health that could have been produced elsewhere with the same resources. India commonly uses approximately ₹2,50,000 per QALY (~1× GDP per capita).

**PSA** (Probabilistic Sensitivity Analysis) — Runs the economic model thousands of times, each time drawing parameter values from their uncertainty distributions, to show the full range of plausible results rather than relying on a single point estimate.
:::

## Part 1: Understanding ICER Basics

The **Incremental Cost-Effectiveness Ratio (ICER)** tells us the additional cost per additional unit of health benefit.

\[
\text{ICER} = \frac{\text{Cost}_{\text{New}} - \text{Cost}_{\text{Standard}}}{\text{Effect}_{\text{New}} - \text{Effect}_{\text{Standard}}}
\]

### Interactive ICER Calculator

```{ojs}
//| panel: sidebar

viewof cost_new = Inputs.range(
  [0, 1000000],
  {value: 200000, step: 5000, label: "Cost of New Treatment (₹):"}
)

viewof cost_std = Inputs.range(
  [0, 1000000],
  {value: 100000, step: 5000, label: "Cost of Standard Treatment (₹):"}
)

viewof qaly_new = Inputs.range(
  [0, 20],
  {value: 8, step: 0.1, label: "QALYs with New Treatment:"}
)

viewof qaly_std = Inputs.range(
  [0, 20],
  {value: 5, step: 0.1, label: "QALYs with Standard Treatment:"}
)

viewof wtp_threshold = Inputs.range(
  [10000, 500000],
  {value: 150000, step: 10000, label: "WTP Threshold (₹/QALY):"}
)
```

```{ojs}
//| panel: fill

// Calculate ICER with all scenarios
incremental_cost = cost_new - cost_std
incremental_qaly = qaly_new - qaly_std

// Determine quadrant
quadrant = {
  if (incremental_cost > 0 && incremental_qaly > 0) return "North-East";
  if (incremental_cost > 0 && incremental_qaly < 0) return "North-West";
  if (incremental_cost < 0 && incremental_qaly > 0) return "South-East";
  if (incremental_cost < 0 && incremental_qaly < 0) return "South-West";
  if (incremental_cost === 0 && incremental_qaly === 0) return "Origin";
  if (incremental_cost === 0) return "Vertical";
  if (incremental_qaly === 0) return "Horizontal";
}

// Calculate ICER with proper handling
icer_value = incremental_qaly !== 0 ? incremental_cost / incremental_qaly : null

// Interpretation logic
interpretation = {
  // Dominant (SE quadrant)
  if (incremental_cost < 0 && incremental_qaly > 0) {
    return {
      status: "Dominant ✓✓",
      color: "darkgreen",
      decision: "Favours adoption",
      explanation: "Less costly AND more effective"
    };
  }

  // Dominated (NW quadrant)
  if (incremental_cost > 0 && incremental_qaly < 0) {
    return {
      status: "Dominated ✗✗",
      color: "darkred",
      decision: "Favours rejection",
      explanation: "More costly AND less effective"
    };
  }

  // SW quadrant - saves money but loses health
  if (incremental_cost < 0 && incremental_qaly < 0) {
    const abs_icer = Math.abs(icer_value);
    const is_acceptable = abs_icer < wtp_threshold;
    return {
      status: is_acceptable ? "Acceptable (trade-off)" : "Not acceptable",
      color: is_acceptable ? "darkorange" : "red",
      decision: is_acceptable ? "May adopt (saves money)" : "Reject (health loss too great)",
      explanation: `Saves ₹${Math.abs(incremental_cost).toLocaleString('en-IN')} but loses ${Math.abs(incremental_qaly).toFixed(2)} QALYs. Society willing to pay ${(abs_icer).toLocaleString('en-IN', {maximumFractionDigits: 0})} per QALY foregone.`
    };
  }

  // NE quadrant - standard ICER comparison
  if (incremental_cost > 0 && incremental_qaly > 0) {
    const is_ce = icer_value < wtp_threshold;
    return {
      status: is_ce ? "Cost-Effective ✓" : "Not Cost-Effective ✗",
      color: is_ce ? "green" : "red",
      decision: is_ce ? "Recommend adoption" : "Do not recommend",
      explanation: `ICER ${is_ce ? 'below' : 'exceeds'} threshold`
    };
  }

  // Edge cases
  if (incremental_cost === 0 && incremental_qaly === 0) {
    return {
      status: "Identical",
      color: "gray",
      decision: "Indifferent",
      explanation: "No difference in costs or effects"
    };
  }

  if (incremental_qaly === 0) {
    return {
      status: incremental_cost > 0 ? "More costly, same effect" : "Less costly, same effect",
      color: incremental_cost > 0 ? "red" : "green",
      decision: incremental_cost > 0 ? "Reject" : "Adopt",
      explanation: "No health benefit difference"
    };
  }

  if (incremental_cost === 0) {
    return {
      status: incremental_qaly > 0 ? "Same cost, more effective" : "Same cost, less effective",
      color: incremental_qaly > 0 ? "green" : "red",
      decision: incremental_qaly > 0 ? "Adopt" : "Reject",
      explanation: "No cost difference"
    };
  }
}

html`
<div style="padding: 20px; background-color: #f8f9fa; border-radius: 8px; margin-bottom: 20px;">
  <h3>Results</h3>
  <table style="width: 100%; border-collapse: collapse; margin-bottom: 15px;">
    <tr>
      <td style="padding: 8px; border: 1px solid #dee2e6; background-color: #e9ecef;"><strong>Incremental Cost:</strong></td>
      <td style="padding: 8px; border: 1px solid #dee2e6;">₹${incremental_cost.toLocaleString('en-IN')}</td>
    </tr>
    <tr>
      <td style="padding: 8px; border: 1px solid #dee2e6; background-color: #e9ecef;"><strong>Incremental QALYs:</strong></td>
      <td style="padding: 8px; border: 1px solid #dee2e6;">${incremental_qaly.toFixed(2)}</td>
    </tr>
    <tr>
      <td style="padding: 8px; border: 1px solid #dee2e6; background-color: #e9ecef;"><strong>CE Plane Quadrant:</strong></td>
      <td style="padding: 8px; border: 1px solid #dee2e6;">${quadrant}</td>
    </tr>
    <tr>
      <td style="padding: 8px; border: 1px solid #dee2e6; background-color: #e9ecef;"><strong>ICER:</strong></td>
      <td style="padding: 8px; border: 1px solid #dee2e6;">
        ${icer_value !== null ? 
          '₹' + icer_value.toLocaleString('en-IN', {maximumFractionDigits: 0}) + '/QALY' : 
          'Undefined (no QALY change)'}
      </td>
    </tr>
  </table>

  <div style="padding: 15px; background-color: white; border-left: 4px solid ${interpretation.color}; border-radius: 4px;">
    <p style="font-size: 20px; font-weight: bold; color: ${interpretation.color}; margin: 0 0 10px 0;">
      ${interpretation.status}
    </p>
    <p style="margin: 5px 0;"><strong>Decision:</strong> ${interpretation.decision}</p>
    <p style="margin: 5px 0; font-style: italic;">${interpretation.explanation}</p>
    <p style="margin: 10px 0 0 0; font-size: 14px; color: #666;">
      WTP threshold: ₹${wtp_threshold.toLocaleString('en-IN')}/QALY
    </p>
  </div>
</div>
`
```

::: {.callout-note}
## Interpretation Guide

### Four Quadrants of CE Plane

**North-East (NE)**: More costly, more effective
- Calculate ICER and compare to threshold
- Most common scenario in HTA

**South-East (SE)**: Less costly, more effective
- **Dominant** — favours adoption
- Best case scenario

**North-West (NW)**: More costly, less effective
- **Dominated** — favours rejection
- Worst case scenario

**South-West (SW)**: Less costly, less effective
- Calculate savings per QALY lost
- Adopt if society values saved resources more than health foregone

### Common Thresholds
- **India**: ~₹2,50,000 per QALY (~1× GDP per capita; used in HTAIn analyses)
- **WHO (older benchmark)**: 1–3× GDP per capita
:::

### Try Different Scenarios

**Example 1 - Dominant:** 
- New: ₹150,000, 8 QALYs
- Standard: ₹200,000, 6 QALYs

**Example 2 - Dominated:**
- New: ₹300,000, 5 QALYs
- Standard: ₹200,000, 7 QALYs

**Example 3 - NE Quadrant:**
- New: ₹250,000, 9 QALYs
- Standard: ₹150,000, 7 QALYs

**Example 4 - SW Quadrant:**
- New: ₹100,000, 7 QALYs
- Standard: ₹180,000, 8.5 QALYs

## Part 2: Cost-Effectiveness Plane

The CE plane helps visualize where an intervention falls relative to standard care and shows uncertainty from probabilistic sensitivity analysis.

```{ojs}
//| panel: sidebar

viewof psa_points = Inputs.range(
  [100, 2000],
  {value: 500, step: 100, label: "PSA Simulations:"}
)

viewof mean_inc_cost = Inputs.range(
  [-200000, 300000],
  {value: 100000, step: 10000, label: "Mean Incremental Cost (₹):"}
)

viewof mean_inc_qaly = Inputs.range(
  [-3, 6],
  {value: 2, step: 0.25, label: "Mean Incremental QALY:"}
)

viewof sd_cost_pct = Inputs.range(
  [10, 80],
  {value: 30, step: 5, label: "Cost Uncertainty (% SD):"}
)

viewof sd_qaly_pct = Inputs.range(
  [10, 80],
  {value: 20, step: 5, label: "QALY Uncertainty (% SD):"}
)

viewof wtp_plane = Inputs.range(
  [50000, 500000],
  {value: 150000, step: 10000, label: "WTP Threshold (₹/QALY):"}
)
```

```{ojs}
//| panel: fill

// Box-Muller transform for normal distribution
function randn_bm() {
  let u = 0, v = 0;
  while(u === 0) u = Math.random();
  while(v === 0) v = Math.random();
  return Math.sqrt(-2.0 * Math.log(u)) * Math.cos(2.0 * Math.PI * v);
}

// Generate PSA simulation data with flexible parameters
psa_data = {
  const sd_cost = Math.abs(mean_inc_cost) * (sd_cost_pct / 100);
  const sd_qaly = Math.abs(mean_inc_qaly) * (sd_qaly_pct / 100);

  return Array.from({length: psa_points}, () => ({
    inc_cost: mean_inc_cost + randn_bm() * sd_cost,
    inc_qaly: mean_inc_qaly + randn_bm() * sd_qaly
  }));
}

// Calculate comprehensive quadrant statistics
quadrant_stats = {
  const ne = psa_data.filter(d => d.inc_cost > 0 && d.inc_qaly > 0).length;
  const nw = psa_data.filter(d => d.inc_cost > 0 && d.inc_qaly < 0).length;
  const se = psa_data.filter(d => d.inc_cost < 0 && d.inc_qaly > 0).length;
  const sw = psa_data.filter(d => d.inc_cost < 0 && d.inc_qaly < 0).length;

  // Cost-effective calculations for NE quadrant
  const ne_ce = psa_data.filter(d => 
    d.inc_cost > 0 && d.inc_qaly > 0 && (d.inc_cost / d.inc_qaly) < wtp_plane
  ).length;

  // Acceptable in SW quadrant (savings justify QALY loss)
  const sw_acceptable = psa_data.filter(d =>
    d.inc_cost < 0 && d.inc_qaly < 0 && Math.abs(d.inc_cost / d.inc_qaly) < wtp_plane
  ).length;

  // Total cost-effective (SE + NE_CE + SW_acceptable)
  const total_ce = se + ne_ce + sw_acceptable;

  return {
    ne: (ne / psa_points * 100).toFixed(1),
    nw: (nw / psa_points * 100).toFixed(1),
    se: (se / psa_points * 100).toFixed(1),
    sw: (sw / psa_points * 100).toFixed(1),
    ne_ce: (ne_ce / psa_points * 100).toFixed(1),
    sw_acceptable: (sw_acceptable / psa_points * 100).toFixed(1),
    prob_ce: (total_ce / psa_points * 100).toFixed(1)
  };
}

// Dynamic axis ranges based on data
axis_ranges = {
  const all_costs = psa_data.map(d => d.inc_cost);
  const all_qalys = psa_data.map(d => d.inc_qaly);

  const min_cost = Math.min(...all_costs, 0);
  const max_cost = Math.max(...all_costs, 0);
  const min_qaly = Math.min(...all_qalys, 0);
  const max_qaly = Math.max(...all_qalys, 0);

  return {
    cost: [min_cost * 1.2, max_cost * 1.2],
    qaly: [min_qaly * 1.2, max_qaly * 1.2]
  };
}

Plot.plot({
  width: 800,
  height: 550,
  grid: true,
  x: {
    label: "Incremental QALYs →",
    domain: axis_ranges.qaly
  },
  y: {
    label: "↑ Incremental Cost (₹)",
    domain: axis_ranges.cost
  },
  marks: [
    // WTP threshold lines (positive and negative slopes)
    Plot.line(
      [
        {x: 0, y: 0}, 
        {x: axis_ranges.qaly[1], y: wtp_plane * axis_ranges.qaly[1]}
      ],
      {x: "x", y: "y", stroke: "red", strokeWidth: 2, strokeDasharray: "5,5"}
    ),
    Plot.line(
      [
        {x: 0, y: 0}, 
        {x: axis_ranges.qaly[0], y: -wtp_plane * axis_ranges.qaly[0]}
      ],
      {x: "x", y: "y", stroke: "red", strokeWidth: 2, strokeDasharray: "5,5"}
    ),

    // Axes
    Plot.ruleX([0], {stroke: "black", strokeWidth: 2}),
    Plot.ruleY([0], {stroke: "black", strokeWidth: 2}),

    // PSA points with color by quadrant
    Plot.dot(psa_data, {
      x: "inc_qaly",
      y: "inc_cost",
      fill: d => {
        if (d.inc_cost < 0 && d.inc_qaly > 0) return "darkgreen"; // SE - dominant
        if (d.inc_cost > 0 && d.inc_qaly < 0) return "darkred";   // NW - dominated
        if (d.inc_cost > 0 && d.inc_qaly > 0) {
          return (d.inc_cost / d.inc_qaly) < wtp_plane ? "steelblue" : "lightcoral"; // NE
        }
        if (d.inc_cost < 0 && d.inc_qaly < 0) {
          return Math.abs(d.inc_cost / d.inc_qaly) < wtp_plane ? "orange" : "pink"; // SW
        }
        return "gray";
      },
      fillOpacity: 0.4,
      r: 3
    }),

    // Mean point
    Plot.dot([{x: mean_inc_qaly, y: mean_inc_cost}], {
      x: "x",
      y: "y",
      fill: "red",
      stroke: "white",
      strokeWidth: 2,
      r: 10,
      symbol: "diamond"
    }),

    // Quadrant labels
    Plot.text([
      {x: axis_ranges.qaly[1] * 0.6, y: axis_ranges.cost[1] * 0.7, 
       label: "NE: More costly,\nmore effective"},
      {x: axis_ranges.qaly[0] * 0.6, y: axis_ranges.cost[1] * 0.7, 
       label: "NW: DOMINATED\n(favours rejection)"},
      {x: axis_ranges.qaly[1] * 0.6, y: axis_ranges.cost[0] * 0.7,
       label: "SE: DOMINANT\n(favours adoption)"},
      {x: axis_ranges.qaly[0] * 0.6, y: axis_ranges.cost[0] * 0.7, 
       label: "SW: Less costly,\nless effective"}
    ], {
      x: "x",
      y: "y",
      text: "label",
      fill: "gray",
      fontSize: 11,
      fontWeight: "bold"
    })
  ]
})

html`
<div style="margin-top: 20px; padding: 15px; background-color: #e7f3ff; border-radius: 8px;">
  <h4>Probabilistic Sensitivity Analysis Results</h4>

  <div style="display: grid; grid-template-columns: 1fr 1fr; gap: 10px; margin: 15px 0;">
    <div style="padding: 10px; background-color: white; border-radius: 4px;">
      <strong>Quadrant Distribution:</strong>
      <ul style="margin: 5px 0; padding-left: 20px;">
        <li>NE (more costly/effective): ${quadrant_stats.ne}%
          <ul><li style="color: steelblue;">Cost-effective in NE: ${quadrant_stats.ne_ce}%</li></ul>
        </li>
        <li style="color: darkred;">NW (dominated): ${quadrant_stats.nw}%</li>
        <li style="color: darkgreen;">SE (dominant): ${quadrant_stats.se}%</li>
        <li>SW (less costly/effective): ${quadrant_stats.sw}%
          <ul><li style="color: orange;">Acceptable in SW: ${quadrant_stats.sw_acceptable}%</li></ul>
        </li>
      </ul>
    </div>

    <div style="padding: 10px; background-color: white; border-radius: 4px;">
      <strong>Decision Summary:</strong>
      <p style="font-size: 18px; font-weight: bold; color: #0066cc; margin: 10px 0;">
        ${quadrant_stats.prob_ce}% probability cost-effective
      </p>
      <p style="font-size: 14px; margin: 5px 0;">
        At threshold: ₹${wtp_plane.toLocaleString('en-IN')}/QALY
      </p>
      <p style="font-size: 13px; color: #666; margin: 10px 0 0 0;">
        <em>Includes: SE (dominant) + NE (below threshold) + SW (acceptable trade-offs)</em>
      </p>
    </div>
  </div>

  <div style="padding: 10px; background-color: #fff3cd; border-radius: 4px; margin-top: 10px;">
    <strong>💡 Interpretation Tips:</strong>
    <ul style="margin: 5px 0; padding-left: 20px; font-size: 14px;">
      <li><strong>Red diamond</strong> = mean (base-case) result</li>
      <li><strong>Red dashed lines</strong> = WTP threshold boundaries</li>
      <li><strong>Dark green points</strong> = dominant (SE quadrant)</li>
      <li><strong>Dark red points</strong> = dominated (NW quadrant)</li>
      <li><strong>Blue/coral points</strong> = NE quadrant (blue = cost-effective)</li>
      <li><strong>Orange/pink points</strong> = SW quadrant (orange = acceptable)</li>
    </ul>
  </div>
</div>
`
```

::: {.callout-important}
## Understanding Uncertainty

- **Wide scatter** = high uncertainty in results → need more research
- **Narrow scatter** = low uncertainty → confident in decision
- **Points crossing quadrants** = parameter uncertainty affects interpretation
- **Points below/above threshold lines** = decision uncertainty

**Decision Rule**: Generally want >50-60% probability cost-effective for adoption
:::

## Part 3: Cost-Effectiveness Acceptability Curve

The CEAC shows the probability an intervention is cost-effective across different WTP thresholds.

```{ojs}
// Generate comprehensive CEAC data
ceac_data = {
  const thresholds = Array.from({length: 40}, (_, i) => 25000 + i * 12500);

  return thresholds.map(threshold => {
    // SE quadrant - always cost-effective (dominant)
    const se_count = psa_data.filter(d => d.inc_cost < 0 && d.inc_qaly > 0).length;

    // NE quadrant - cost-effective if ICER < threshold
    const ne_ce = psa_data.filter(d => 
      d.inc_cost > 0 && d.inc_qaly > 0 && (d.inc_cost / d.inc_qaly) < threshold
    ).length;

    // SW quadrant - acceptable if savings justify QALY loss
    const sw_accept = psa_data.filter(d =>
      d.inc_cost < 0 && d.inc_qaly < 0 && Math.abs(d.inc_cost / d.inc_qaly) < threshold
    ).length;

    // NW quadrant - never acceptable (dominated)

    const total_acceptable = se_count + ne_ce + sw_accept;

    return {
      threshold: threshold,
      probability: total_acceptable / psa_points,
      se_prob: se_count / psa_points,
      ne_prob: ne_ce / psa_points,
      sw_prob: sw_accept / psa_points
    };
  });
}

Plot.plot({
  width: 800,
  height: 450,
  marginLeft: 70,
  marginBottom: 60,
  x: {
    label: "Willingness-to-Pay Threshold (₹/QALY) →",
    tickFormat: d => `₹${(d/1000).toFixed(0)}k`
  },
  y: {
    label: "↑ Probability Cost-Effective",
    domain: [0, 1],
    tickFormat: d => `${(d * 100).toFixed(0)}%`
  },
  grid: true,
  marks: [
    // Total CEAC line
    Plot.line(ceac_data, {
      x: "threshold",
      y: "probability",
      stroke: "#0066cc",
      strokeWidth: 4
    }),
    Plot.dot(ceac_data, {
      x: "threshold",
      y: "probability",
      fill: "#0066cc",
      r: 3
    }),

    // Current threshold line
    Plot.ruleX([wtp_plane], {
      stroke: "red",
      strokeWidth: 2,
      strokeDasharray: "5,5"
    }),

    // 50% probability line
    Plot.ruleY([0.5], {
      stroke: "gray",
      strokeWidth: 1,
      strokeDasharray: "3,3"
    }),

    // Indian threshold range shading
    Plot.rectY([{x1: 100000, x2: 150000, y: 1}], {
      x1: "x1",
      x2: "x2",
      y: "y",
      fill: "lightgreen",
      fillOpacity: 0.1
    })
  ]
})

// Find probability at selected threshold
prob_at_threshold = ceac_data.find(d => Math.abs(d.threshold - wtp_plane) < 13000)

html`
<div style="margin-top: 15px;">
  <div style="padding: 15px; background-color: #fff3cd; border-radius: 8px; margin-bottom: 10px;">
    <p><strong>At selected threshold of ₹${wtp_plane.toLocaleString('en-IN')}/QALY:</strong></p>
    <p style="font-size: 22px; font-weight: bold; color: #856404; margin: 10px 0;">
      ${prob_at_threshold ? (prob_at_threshold.probability * 100).toFixed(1) : 'N/A'}% probability cost-effective
    </p>
    ${prob_at_threshold ? `
      <div style="font-size: 14px; color: #666;">
        <p style="margin: 5px 0;">Breakdown by quadrant:</p>
        <ul style="margin: 5px 0; padding-left: 20px;">
          <li>SE (dominant): ${(prob_at_threshold.se_prob * 100).toFixed(1)}%</li>
          <li>NE (cost-effective): ${(prob_at_threshold.ne_prob * 100).toFixed(1)}%</li>
          <li>SW (acceptable): ${(prob_at_threshold.sw_prob * 100).toFixed(1)}%</li>
        </ul>
      </div>
    ` : ''}
  </div>

  <div style="padding: 15px; background-color: #d1ecf1; border-radius: 8px;">
    <strong>📊 CEAC Interpretation:</strong>
    <ul style="margin: 5px 0; padding-left: 20px; font-size: 14px;">
      <li><strong>Steep curve</strong>: Result highly sensitive to threshold choice</li>
      <li><strong>Flat curve</strong>: Result robust across thresholds</li>
      <li><strong>Green shaded area</strong>: Common Indian WTP range (₹1-1.5 lakh/QALY)</li>
      <li><strong>Above 50% line</strong>: More likely cost-effective than not</li>
      <li><strong>Curve never reaches 100%</strong>: Some simulations always in NW (dominated)</li>
    </ul>
  </div>
</div>
`
```

::: {.callout-note}
## How to Read CEAC

The CEAC answers: "What's the probability this intervention is cost-effective at different thresholds?"

**Key Features:**
- **X-axis**: Different WTP threshold values
- **Y-axis**: Probability of being cost-effective
- **Steep sections**: Uncertainty concentrated in that threshold range
- **Plateau regions**: Probability stable (less sensitivity to threshold)

**Decision Guidelines:**
- **>70-80%**: Strong evidence for adoption
- **50-70%**: Moderate evidence, may need additional research
- **<50%**: Weak evidence, likely not cost-effective
:::

## Part 4: Deterministic Sensitivity Analysis (Tornado Diagram)

One-way (deterministic) sensitivity analysis varies each parameter individually while holding others at base-case values. The **tornado diagram** ranks parameters by their influence on the result.

```{r}
library(tidyverse)
set.seed(123)

n <- 5000
delta_qaly <- rnorm(n, mean=0.04, sd=0.03)
delta_cost <- rnorm(n, mean=6000, sd=7000)
psa <- tibble(delta_cost, delta_qaly)

lambda <- 250000
base_nmb <- lambda * mean(psa$delta_qaly) - mean(psa$delta_cost)

params <- tibble(
  param = c("Treatment cost", "Utility gain", "Hospitalisation rate", "Adherence", "Mortality risk"),
  low  = c(-12000, -9000, -7000, -5000, -4000),
  high = c( 14000,  11000,  9000,  7000,  6000)
) |>
  mutate(low_nmb  = base_nmb + low,
         high_nmb = base_nmb + high)

params_long <- params |>
  pivot_longer(cols = c(low_nmb, high_nmb), names_to = "type", values_to = "nmb")

ggplot(params_long, aes(nmb, reorder(param, nmb), fill = type)) +
  geom_col(position = "identity", alpha = .8) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Net Monetary Benefit (Rs.)",
       y = "Parameter",
       title = "Tornado Diagram") +
  theme_minimal(base_size = 16) +
  theme(legend.position = "none")
```

::: {.callout-note}
## Interpreting the Tornado Diagram

- **Long bars** indicate parameters that strongly influence the decision — these are priorities for better evidence.
- **Short bars** indicate parameters where uncertainty matters less.
- If a bar **crosses zero**, varying that single parameter can reverse the adoption decision.
- The tornado diagram complements PSA: it shows *which* parameters drive uncertainty, while PSA shows *overall* decision uncertainty.
:::

## Part 5: Expected Value of Perfect Information (EVPI)

EVPI quantifies the maximum amount a decision-maker should be willing to pay for additional research to eliminate all parameter uncertainty before making a decision.

```{r}
lambda_seq <- seq(0, 300000, 5000)

evpi <- map_df(lambda_seq, function(l) {
  nmb <- l * psa$delta_qaly - psa$delta_cost
  evpi_val <- mean(pmax(nmb, 0)) - max(mean(nmb), 0)
  tibble(wtp = l, evpi = evpi_val)
})

ggplot(evpi, aes(wtp, evpi)) +
  geom_line(linewidth = 1.4, colour = "#d95f02") +
  geom_vline(xintercept = 250000, linetype = "dashed", colour = "grey50") +
  labs(x = "WTP (Rs./QALY)",
       y = "EVPI per patient (Rs.)",
       title = "Expected Value of Perfect Information") +
  theme_minimal(base_size = 16)
```

::: {.callout-note}
## Interpreting EVPI

- **Low EVPI** at the policy threshold suggests current evidence is sufficient — proceed with the decision.
- **High EVPI** suggests substantial value in commissioning further research before committing resources.
- EVPI peaks near thresholds where the decision is most uncertain (probability of cost-effectiveness close to 50%).
- EVPI is always expressed **per patient**; multiply by the eligible population to estimate the total value of research.
:::

## Part 6: Budget Impact Analysis

Cost-effectiveness does not guarantee affordability. Budget impact analysis estimates the financial consequences of adopting a new intervention at the health-system level over a defined time horizon.

```{r}
years <- 1:5
uptake <- c(.10, .25, .45, .65, .80)
eligible <- 200000
cost_patient <- 1500

budget <- tibble(
  year = years,
  treated = eligible * uptake,
  annual_cost = treated * cost_patient
)

ggplot(budget, aes(year, annual_cost / 1e7)) +
  geom_col(fill = "#54278f") +
  labs(x = "Year",
       y = "Budget Impact (Crores Rs.)",
       title = "5-Year Budget Impact with Gradual Uptake") +
  theme_minimal(base_size = 16)
```

::: {.callout-note}
## Interpreting Budget Impact

- Budget impact depends on **eligible population size** and **uptake trajectory**, not just per-patient cost.
- An intervention can be cost-effective (favourable ICER) yet **unaffordable** if the eligible population is large.
- Decision-makers use BIA alongside CE results to assess whether the health system can absorb the expenditure within existing or projected budgets.
:::

## Part 7: Real-World Case Study

**Scenario:** Portable Peritoneal Dialysis vs. In-Center Hemodialysis for End-Stage Renal Disease in India

```{ojs}
// Real case data
case_data = ({
  intervention: "Portable Peritoneal Dialysis",
  comparator: "In-Center Hemodialysis",
  cost_int: 185000,
  cost_comp: 250000,
  qaly_int: 6.2,
  qaly_comp: 5.8,
  inc_cost: -65000,
  inc_qaly: 0.4,
  icer: -162500
})

html`
<div style="padding: 20px; background-color: #f8f9fa; border-radius: 8px; margin: 20px 0;">
  <h4>Base Case Results</h4>
  <table style="width: 100%; border-collapse: collapse;">
    <thead>
      <tr style="background-color: #e9ecef;">
        <th style="padding: 10px; text-align: left; border: 1px solid #dee2e6;">Strategy</th>
        <th style="padding: 10px; text-align: right; border: 1px solid #dee2e6;">Total Cost (₹)</th>
        <th style="padding: 10px; text-align: right; border: 1px solid #dee2e6;">Total QALYs</th>
      </tr>
    </thead>
    <tbody>
      <tr>
        <td style="padding: 10px; border: 1px solid #dee2e6;">${case_data.intervention}</td>
        <td style="padding: 10px; text-align: right; border: 1px solid #dee2e6;">₹${case_data.cost_int.toLocaleString('en-IN')}</td>
        <td style="padding: 10px; text-align: right; border: 1px solid #dee2e6;">${case_data.qaly_int.toFixed(2)}</td>
      </tr>
      <tr>
        <td style="padding: 10px; border: 1px solid #dee2e6;">${case_data.comparator}</td>
        <td style="padding: 10px; text-align: right; border: 1px solid #dee2e6;">₹${case_data.cost_comp.toLocaleString('en-IN')}</td>
        <td style="padding: 10px; text-align: right; border: 1px solid #dee2e6;">${case_data.qaly_comp.toFixed(2)}</td>
      </tr>
      <tr style="background-color: #d1ecf1; font-weight: bold;">
        <td style="padding: 10px; border: 1px solid #dee2e6;">Incremental</td>
        <td style="padding: 10px; text-align: right; border: 1px solid #dee2e6; color: green;">₹${case_data.inc_cost.toLocaleString('en-IN')}</td>
        <td style="padding: 10px; text-align: right; border: 1px solid #dee2e6; color: green;">${case_data.inc_qaly.toFixed(2)}</td>
      </tr>
    </tbody>
  </table>
  <div style="margin-top: 15px; padding: 15px; background-color: #d4edda; border-radius: 4px; border-left: 4px solid darkgreen;">
    <p style="font-size: 20px; font-weight: bold; color: darkgreen; margin: 0 0 5px 0;">
      Result: DOMINANT ✓✓
    </p>
    <p style="margin: 5px 0;">Portable dialysis is less costly AND more effective</p>
    <p style="margin: 5px 0; font-size: 14px; color: #155724;">
      <strong>Location:</strong> South-East quadrant of CE plane<br>
      <strong>Decision:</strong> Strong case for adoption<br>
      <strong>ICER:</strong> ₹${case_data.icer.toLocaleString('en-IN')}/QALY (saves money, gains health)
    </p>
  </div>
</div>
`
```

### Group Exercise

Answer these questions based on the case above:

```{ojs}
viewof q1_answer = Inputs.radio(
  ["Cost-effective", "Not cost-effective", "Dominant", "Dominated"],
  {label: "Q1: What is the cost-effectiveness status?", value: null}
)

viewof q2_answer = Inputs.radio(
  ["North-East", "North-West", "South-East", "South-West"],
  {label: "Q2: Which quadrant of CE plane?", value: null}
)

viewof q3_answer = Inputs.radio(
  ["Strong case for adoption", "Recommend with conditions", "Do not recommend", "Need more data"],
  {label: "Q3: Policy recommendation?", value: null}
)

viewof q4_answer = Inputs.radio(
  ["Yes, always", "No, never", "Only if below threshold", "Depends on uncertainty"],
  {label: "Q4: Do we need to compare ICER to threshold?", value: null}
)

// Button with proper handling
viewof show_answers_btn = Inputs.button("Show Correct Answers")
```

```{ojs}
// Track button clicks
show_answers_count = show_answers_btn

html`
${show_answers_count > 0 ? `
<div style="margin-top: 20px; padding: 15px; background-color: #d4edda; border-radius: 8px;">
  <h4>✓ Correct Answers:</h4>
  <p><strong>Q1:</strong> Dominant (provides more QALYs at lower cost)</p>
  <p><strong>Q2:</strong> South-East quadrant (less costly, more effective)</p>
  <p><strong>Q3:</strong> Strong case for adoption (subject to equity, feasibility, and budget considerations)</p>
  <p><strong>Q4:</strong> No, never - Dominant strategies don't require threshold comparison</p>

  <div style="margin-top: 15px; padding: 10px; background-color: white; border-radius: 4px;">
    <p style="font-weight: bold; color: #155724;">💡 Key Learning Points:</p>
    <ul style="margin: 5px 0; padding-left: 20px;">
      <li>Dominant interventions (SE quadrant) have the <strong>strongest CE case</strong> for adoption</li>
      <li>No need to compare ICER to threshold for dominant strategies</li>
      <li>This is the ideal CE outcome — better health at lower cost</li>
      <li>Final decisions still consider equity, budget feasibility, and implementation context</li>
    </ul>
  </div>
</div>
` : ''}
`
```

## Part 8: Spot the Error

Identify mistakes in these common misinterpretations:

```{ojs}
viewof scenario = Inputs.select(
  ["Scenario 1", "Scenario 2", "Scenario 3", "Scenario 4", "Scenario 5"],
  {label: "Select scenario:", value: "Scenario 1"}
)

scenarios = ({
  "Scenario 1": {
    statement: "The ICER is ₹180,000/QALY. Since this is positive, the intervention is cost-effective.",
    error: "Confusing positive ICER with cost-effectiveness. Must compare to WTP threshold.",
    correct: "An ICER of ₹180,000/QALY exceeds typical Indian thresholds (₹100k-150k), so likely NOT cost-effective. Positive ICER only means more costly - doesn't indicate value for money."
  },
  "Scenario 2": {
    statement: "The intervention is in the North-West quadrant, but has a low ICER, so we should adopt it.",
    error: "North-West = more costly AND less effective (dominated). Should NEVER be adopted regardless of ICER value.",
    correct: "North-West quadrant interventions are dominated and should be rejected immediately. No ICER comparison needed - it's strictly worse than the comparator."
  },
  "Scenario 3": {
    statement: "The base-case ICER is ₹120,000/QALY (below threshold), so the intervention is cost-effective.",
    error: "Ignoring uncertainty. Must consider PSA results and probability of cost-effectiveness.",
    correct: "While base-case looks favorable, check PSA scatter and CEAC. If probability <50%, there's high decision uncertainty. May need more research."
  },
  "Scenario 4": {
    statement: "The ICER is negative (−₹50,000/QALY), which means the intervention is not cost-effective.",
    error: "Misunderstanding negative ICERs. Could be dominant (SE quadrant) or in SW quadrant - need more information.",
    correct: "Negative ICER needs context. If in SE quadrant (saves money, gains QALYs) = dominant, favours adoption. If in SW quadrant (saves money, loses QALYs) = compare absolute value to threshold."
  },
  "Scenario 5": {
    statement: "The intervention is in the SW quadrant (saves ₹80,000, loses 0.4 QALYs). Since it saves money, we should adopt it.",
    error: "Ignoring health losses. Must calculate whether society values the cost savings more than the QALYs foregone.",
    correct: "ICER = −₹200,000/QALY (society pays ₹200,000 per QALY foregone). If threshold is ₹150,000/QALY, this exceeds willingness-to-pay for health, so reject. The cost savings don't justify the health loss."
  }
})

html`
<div style="padding: 20px; background-color: #fff3cd; border-radius: 8px; margin: 20px 0;">
  <h4>Incorrect Interpretation:</h4>
  <p style="font-size: 16px; font-style: italic; padding: 15px; background-color: white; border-left: 4px solid #ffc107; margin: 10px 0;">
    "${scenarios[scenario].statement}"
  </p>
  <button id="reveal-${scenario}" onclick="
    document.getElementById('error-${scenario}').style.display = 'block';
    document.getElementById('correct-${scenario}').style.display = 'block';
    this.style.display = 'none';
  " style="margin-top: 10px; padding: 8px 16px; background-color: #007bff; color: white; border: none; border-radius: 4px; cursor: pointer;">
    Reveal Error & Correction
  </button>
  <div id="error-${scenario}" style="display: none; margin-top: 15px; padding: 15px; background-color: #f8d7da; border-radius: 4px; border-left: 4px solid #dc3545;">
    <p style="margin: 0;"><strong>❌ Error:</strong> ${scenarios[scenario].error}</p>
  </div>
  <div id="correct-${scenario}" style="display: none; margin-top: 10px; padding: 15px; background-color: #d4edda; border-radius: 4px; border-left: 4px solid #28a745;">
    <p style="margin: 0;"><strong>✓ Correct Interpretation:</strong> ${scenarios[scenario].correct}</p>
  </div>
</div>
`
```

## Takeaway: HTA Interpretation Checklist

Use this systematic framework for interpreting any economic evaluation:

### 1. Base-Case Results
- [ ] Calculate incremental costs and effects
- [ ] Determine CE plane quadrant (NE, NW, SE, SW)
- [ ] Calculate ICER (if applicable)
- [ ] Compare to appropriate WTP threshold
- [ ] Check for dominance or dominated strategies

### 2. Uncertainty Analysis
- [ ] Review PSA scatter plot distribution across quadrants
- [ ] Check CEAC at relevant thresholds
- [ ] Assess probability of cost-effectiveness (target >50-70%)
- [ ] Evaluate robustness of findings

### 3. Deterministic Sensitivity Analysis
- [ ] Review tornado diagram for key drivers
- [ ] Check whether any single parameter reverses the decision (bar crosses zero)
- [ ] Prioritise parameters with largest influence for further evidence

### 4. Research Value
- [ ] Check EVPI at the policy threshold
- [ ] If EVPI is high, consider whether further research is warranted before deciding
- [ ] Use EVPPI (if available) to identify which parameter merits targeted research

### 5. Budget Impact & Affordability
- [ ] Estimate eligible population and likely uptake trajectory
- [ ] Project annual and cumulative budget impact over 3–5 years
- [ ] Assess whether cost-effective intervention is also affordable within current budgets

### 6. Context & Applicability
- [ ] Are costs relevant to Indian healthcare setting?
- [ ] Is the comparator appropriate for local clinical practice?
- [ ] Are outcomes meaningful to patients and policymakers?
- [ ] Consider equity and access implications

### 7. Decision Recommendation
- [ ] State clear cost-effectiveness status
- [ ] Express confidence level (based on PSA/CEAC)
- [ ] List any conditions or caveats
- [ ] Identify research gaps or areas of uncertainty
- [ ] Consider implementation and feasibility factors

::: {.callout-tip}
## Continue Practicing
- [Practice Cases with Solutions](practice.qmd)
- [Additional Resources](resources.qmd)
:::

---

## Next Steps

- [Practice with more cases](practice.qmd)
- [Recommended reading and resources](resources.qmd)

**Questions?** Contact [hta@aiimsbhopal.edu.in](mailto:hta@aiimsbhopal.edu.in)
