---
title: "Interpretation of Economic Evaluation Results in HTA"
subtitle: "How Decision Makers Actually Read Cost-Effectiveness Studies"
author: "RRC-HTA, AIIMS Bhopal"
format:
  revealjs:
    theme: simple
    slide-number: c/t
    transition: fade
    center: false
    footer: "HTA Workshop | IGGMC Nagpur | Feb 24-25, 2026"
    css: styles_slides.css
    math: mathjax
    code-overflow: wrap
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(scales)
set.seed(123)
```

## Why This Session Exists

Clinicians usually ask:

> "Is the ICER below the threshold?"

Policy makers instead ask:

> "If we fund this — what health do we lose elsewhere?"

Health Technology Assessment is a **resource allocation decision under a fixed health budget**.

---

## Learning Objectives

By the end of this session, you will be able to:

1. Interpret ICER results correctly across all quadrants
2. Understand dominance and extended dominance
3. Read the cost-effectiveness plane and its quadrants
4. Interpret CEAC probabilities for decision-making
5. Use Net Monetary Benefit as a decision rule
6. Judge uncertainty and research value (EVPI/EVPPI)
7. Translate results into policy recommendations

---

## Quick Recap: Key Concepts

*(Covered in detail by earlier sessions — brief recap for context)*

**QALY** (Quality-Adjusted Life Year): One year of life in perfect health = 1 QALY. Captures both length and quality of life. A utility of 0.7 for 2 years = 1.4 QALYs.

**WTP threshold**: The maximum a health system is willing to pay per QALY gained. Reflects opportunity cost — health that could be produced elsewhere with the same money.

**PSA** (Probabilistic Sensitivity Analysis): Runs the model thousands of times, each time sampling parameters from their uncertainty distributions, to show the range of plausible results — not just one point estimate.

---

## Incremental Analysis

$$
ICER = \frac{\Delta C}{\Delta E} = \frac{C_{new}-C_{current}}{E_{new}-E_{current}}
$$

Additional rupees required to generate one additional QALY relative to current care.

---

## Net Monetary Benefit

$$
NMB(\lambda) = \lambda \times \Delta E - \Delta C
$$

Adopt if **NMB > 0**

NMB avoids the ratio problems of ICER — no division by zero, no ambiguous negative values.

---

## Opportunity Cost and the Threshold

A threshold does **not** represent a "price society likes".

It represents **health produced elsewhere** in the system using the same resources.

Funding above this level causes **net population health loss**.

::: {.callout-note}
India's commonly used threshold: **₹2,50,000 per QALY** (~1× GDP per capita)
:::

---

## Simulated Economic Evaluation

```{r}
n <- 5000
delta_qaly <- rnorm(n, mean=0.04, sd=0.03)
delta_cost <- rnorm(n, mean=6000, sd=7000)
psa <- tibble(delta_cost, delta_qaly)
```

We simulate 5,000 PSA iterations for a hypothetical intervention:

- Mean incremental QALY: **0.04**
- Mean incremental cost: **₹6,000**
- Both with considerable uncertainty

---

## Cost-Effectiveness Plane

```{r}
wtp <- 250000

mean_pt <- psa |> summarise(dc=mean(delta_cost), de=mean(delta_qaly))
ce_prob <- mean(wtp*psa$delta_qaly - psa$delta_cost > 0)

ggplot(psa, aes(delta_qaly, delta_cost))+
  geom_point(alpha=.25, colour="#2b8cbe", size=1.5)+
  geom_hline(yintercept=0, linewidth=.5)+
  geom_vline(xintercept=0, linewidth=.5)+
  geom_abline(intercept=0, slope=wtp, colour="red", linewidth=1.2)+
  geom_point(data=mean_pt, aes(de,dc), size=5, colour="black")+
  annotate("text", x=.12, y=15000,
           label=paste0(round(ce_prob*100,1),"% cost-effective"),
           size=6)+
  annotate("text", x=.10, y=20000, label="NE", size=5, fontface="bold")+
  annotate("text", x=-.06, y=20000, label="NW", size=5, fontface="bold")+
  annotate("text", x=.10, y=-15000, label="SE", size=5, fontface="bold")+
  annotate("text", x=-.06, y=-15000, label="SW", size=5, fontface="bold")+
  labs(x="Incremental QALYs",
       y="Incremental Costs (Rs.)",
       title="Cost-Effectiveness Plane")+
  theme_minimal(base_size=18)
```

---

## Four Quadrants of the CE Plane

| Quadrant | Cost | Effect | Decision |
|:---------|:-----|:-------|:---------|
| **NE** | Higher | Better | Compare ICER to threshold |
| **SE** | Lower | Better | **Dominant** — favours adoption |
| **NW** | Higher | Worse | **Dominated** — favours rejection |
| **SW** | Lower | Worse | Compare savings to QALY loss |

::: {.callout-important}
Only NE and SW quadrants require threshold comparison. SE and NW provide strong *prima facie* guidance, but real-world decisions also weigh equity, feasibility, and local context.
:::

---

## How to Read the Cost-Effectiveness Plane

* Each dot = one possible real-world outcome from PSA
* Red line = WTP decision threshold (₹2,50,000/QALY)
* **Below** the line → acceptable value for money
* **Above** the line → poor value for money

Probability of cost-effectiveness = proportion of PSA points below the threshold line.

---

## Cost-Effectiveness Acceptability Curve

```{r}
wtp_seq <- seq(0,300000,5000)

ceac <- map_df(wtp_seq, function(l){
  tibble(
    wtp=l,
    prob=mean(l*psa$delta_qaly - psa$delta_cost > 0)
  )
})

ggplot(ceac, aes(wtp, prob))+
  geom_line(linewidth=1.4, colour="#2c7fb8")+
  geom_vline(xintercept=250000, linetype="dashed", linewidth=1)+
  geom_point(data=filter(ceac, wtp==250000), size=4, colour="red")+
  annotate("rect", xmin=0, xmax=250000, ymin=0, ymax=1,
           alpha=.08, fill="green")+
  scale_y_continuous(labels=scales::percent, limits=c(0,1))+
  labs(x="Willingness-to-Pay (Rs. per QALY)",
       y="Probability Cost-Effective",
       title="Cost-Effectiveness Acceptability Curve")+
  theme_minimal(base_size=18)
```

---

## How to Read the CEAC

* X-axis → willingness to pay threshold
* Y-axis → probability the adoption decision is correct

At the Indian policy threshold (₹2,50,000/QALY), the CEAC gives the **confidence level** for adoption.

**Decision guidelines:**

* >70–80%: Strong CE evidence — supports adoption
* 50–70%: Moderate CE evidence — consider additional research
* <50%: Weak CE evidence — adoption hard to justify on CE grounds alone

---

## Deterministic Sensitivity Analysis

```{r}
lambda <- 250000
base_nmb <- lambda*mean(psa$delta_qaly) - mean(psa$delta_cost)

params <- tibble(
  param=c("Treatment cost","Utility gain","Hospitalisation rate","Adherence","Mortality risk"),
  low=c(-12000,-9000,-7000,-5000,-4000),
  high=c(14000,11000,9000,7000,6000)
)

params <- params |>
  mutate(low_nmb = base_nmb + low,
         high_nmb = base_nmb + high)

params_long <- params |>
  pivot_longer(cols=c(low_nmb,high_nmb), names_to="type", values_to="nmb")

ggplot(params_long, aes(nmb, reorder(param, nmb), fill=type))+
  geom_col(position="identity", alpha=.8)+
  geom_vline(xintercept=0, linetype="dashed")+
  labs(x="Net Monetary Benefit (Rs.)",
       y="Parameter",
       title="Tornado Diagram")+
  theme_minimal(base_size=18)+
  theme(legend.position="none")
```

---

## How to Interpret the Tornado Diagram

* **Long bars** → influential parameters (drive the decision)
* **Short bars** → unimportant assumptions

The **longest bar** identifies what evidence is needed next.

If a bar **crosses zero**, that parameter alone can reverse the decision.

---

## Efficiency Frontier

```{r}
interv <- tibble(
  name=c("Current care","A","B","C"),
  cost=c(20000,26000,34000,48000),
  qaly=c(6.0,6.5,6.9,7.1)
)

ggplot(interv, aes(qaly, cost, label=name))+
  geom_point(size=4)+
  geom_line(linewidth=1, colour="#2c7fb8")+
  geom_text(nudge_y=2000, size=5)+
  labs(x="QALY",
       y="Cost (Rs.)",
       title="Cost-Effectiveness Frontier")+
  theme_minimal(base_size=18)
```

The frontier connects **non-dominated strategies**. Options below the frontier are dominated or extendedly dominated.

---

## Expected Value of Perfect Information (EVPI)

```{r}
lambda_seq <- seq(0,300000,5000)

evpi <- map_df(lambda_seq, function(l){
  nmb <- l*psa$delta_qaly - psa$delta_cost
  evpi_val <- mean(pmax(nmb,0)) - max(mean(nmb),0)
  tibble(wtp=l, evpi=evpi_val)
})

ggplot(evpi, aes(wtp,evpi))+
  geom_line(linewidth=1.4, colour="#d95f02")+
  geom_vline(xintercept=250000, linetype="dashed", colour="grey50")+
  labs(x="WTP (Rs./QALY)",
       y="EVPI per patient (Rs.)",
       title="Expected Value of Perfect Information")+
  theme_minimal(base_size=18)
```

---

## What EVPI and EVPPI Mean

**EVPI** answers: Is additional research worth funding?

* Low EVPI → **decide now** with current evidence
* High EVPI → **commission research** before deciding

**EVPPI** goes further — it identifies **which parameter** merits further research.

```{r}
utility <- rnorm(n, 0.04, 0.03)
cost_sim <- psa$delta_cost
lambda_evppi <- 250000

nmb <- lambda_evppi * utility - cost_sim

evppi <- mean(tapply(nmb, cut(utility, 10), function(x) max(mean(x), 0))) -
         max(mean(nmb), 0)

tibble(Parameter = "Utility gain", `EVPPI (Rs.)` = round(evppi, 0)) |>
  knitr::kable()
```

High EVPPI for a parameter → prioritise research to reduce that uncertainty.

---

## Budget Impact

```{r}
years <- 1:5
uptake <- c(.1,.25,.45,.65,.8)
eligible <- 200000
cost_patient <- 1500

budget <- tibble(
  year=years,
  treated=eligible*uptake,
  annual_cost=treated*cost_patient
)

ggplot(budget,aes(year,annual_cost/1e7))+
  geom_col(fill="#54278f")+
  labs(x="Year",
       y="Budget Impact (Crores Rs.)",
       title="5-Year Budget Impact with Uptake")+
  theme_minimal(base_size=18)
```

Cost-effectiveness ≠ affordability. Budget impact analysis answers: **Can we actually pay for this?**

---

## Real-World Case: Portable Dialysis vs Hemodialysis

| | Portable PD | In-Centre HD |
|:--|:--|:--|
| **Annual cost** | ₹1,85,000 | ₹2,50,000 |
| **QALYs** | 6.2 | 5.8 |

**Incremental:** Saves ₹65,000 and gains 0.4 QALYs

::: {.callout-tip}
**Result: DOMINANT** (SE quadrant). Less costly AND more effective. No threshold comparison needed — **strong case for adoption**, subject to feasibility and equity considerations.
:::

---

## Common Interpretation Pitfalls

::: {.incremental}

1. **"Positive ICER = cost-effective"** — Wrong. Must compare to threshold.

2. **"NW quadrant, low ICER — adopt it"** — Wrong. NW = dominated. No CE basis for adoption.

3. **"Base-case below threshold → done"** — Wrong. Must check PSA uncertainty.

4. **"Negative ICER = bad"** — Wrong. Could be dominant (SE) or a SW trade-off.

5. **"Saves money, so adopt"** (SW) — Wrong. Must check if QALY loss is acceptable.

:::

---

## HTA Interpretation Checklist

**1. Base-Case Results**

- Calculate incremental costs and effects
- Determine CE plane quadrant
- Calculate ICER (if NE or SW)
- Compare to WTP threshold

**2. Uncertainty**

- Review PSA scatter and CEAC
- Identify key drivers (tornado)
- Assess probability of cost-effectiveness

**3. Context**

- Are costs relevant to Indian setting?
- Assess budget impact and affordability
- Consider equity and access

---

## Final Interpretation

HTA decisions combine:

* Cost-effectiveness (NMB)
* Uncertainty (CEAC)
* Drivers (tornado)
* Research value (EVPI/EVPPI)
* Affordability (budget impact)

**ICER is a statistic.**
**HTA is a deliberative decision** — informed by evidence, shaped by values, equity, and context.

---

## Thank You

::: {.columns}
::: {.column width="60%"}
Regional Resource Centre for Health Technology Assessment (RRC-HTA)

All India Institute of Medical Sciences (AIIMS), Bhopal

Email: [hta@aiimsbhopal.edu.in](mailto:hta@aiimsbhopal.edu.in)
:::

::: {.column width="40%"}
**Continue learning:**

- [Interactive Session](session.html)
- [Practice Cases](practice.html)
- [Resources](resources.html)
:::
:::
